Estimation of Stochastic Wilson and Cowan Neural Field Model using Expectation Maximisation Algorithm

### 2011-08-10 ltx/data_paper/DataPaper.tex
edits to results discussion and the conclusion

@@

This section demonstrates the identification of the proposed IDE model using simulated data. The observation noise and the field disturbance were unknown to the estimation algorithm and no assumption were made on the shape of the connectivity kernel. Data was generated employing state-space equation (\ref{eq:finalformstatespacemodel}-\ref{eq:QmatrixForSigmapoints}) using a square grid of $n_x= 9 \times 9$ equally spaced ($2.5~mm$) basis functions with $\sigma_{\phi}^2=1.58~mm^2$, assuming the observed field can be fully described by the decomposed IDE model. 100 realizations of 500 ms of data were generated and the first 100 ms was discarded allowing the model's dynamic to reach the steady state. Observation noise variance was set to 0.2. All the other parameters for the simulation were chosen as that of Chapter 2 (see Table~\ref{tab:Parameters}).

The correlation analysis described in \S\ref{subsec:CorrAnal} were applied to the generated data in order to approximate the unknown supports of both the connectivity kernel and disturbance covariance function. In doing so, the unknown parameters $\sigma_{\epsilon}^2$ and $\xi$ are required and should be guessed beforehand.  To form a more accurate initial guess of the unknown parameters and explore the effect of the chosen initial guess on the estimated supports several tests were performed. First consider  \eqref{eq:EM-MMGFourier2} for estimating the disturbance covariance function. As mentioned before the power spectral densities are non-negative, therefore the minimum value of $\mathcal F \{\mathbf E\left[ R_{y_ty_t}\right] \}$ and $\mathcal F \{\mathbf E\left[ R_{y_{t+1}y_{t+1}}\right]\}$ should be chosen as the higher limit for the initial guess of $\sigma^2_{\varepsilon}$. The minimum value of the power spectral density was found to be 0.23, Figure~\ref{fig:DisturbanceSupportExperiment} shows the support estimates for varying  $\sigma^2_{\varepsilon}$ around the calculated limit. This figure shows that for values less than 0.23 the estimated support is close to the true one (see Figure~\ref{fig:BinaryDisturbanceWidthEstimation}(b)) while for values above 0.23 the shape of the disturbance covariance function is distorted. The estimated support of the covariance function convolved twice with the sensor kernel for  $\sigma^2_{\varepsilon}<$0.23 is shown in Figure~\ref{fig:BinaryDisturbanceWidthEstimation} (a), where the results were normalised and threshold for values less than 0.01. By fitting a Gaussian basis function to the result, and under the assumption that the width of the sensor, $\sigma_m^2$ were known, the width of the disturbance covariance function, $\sigma_d^2$ was analytically calculated  using \eqref{eq:EM-AnalyticSupportCalculation}. Alternatively deconvolution can be performed numerically to directly find the support as shown in Figure~\ref{fig:DisturbanceWidthEstimationCrossSection}.%Figure 1 here

@@

The effect of the observation noise variance on the connectivity kernel support is very much like to that of the disturbance covariance function where an initial guess can be made in a similar manner. To study the influence of the synaptic parameter, $\xi=1-\frac{ T_s}{\tau}$, on the resultant support of the connectivity kernel different values of synaptic time constant, $\tau$, in a physiologically plausible range were considered  while noting that the chosen values should be higher than the sampling time, $T_s$. In fact synaptic parameter in \eqref{eq:EM-Fourier_TF_of_Kernel}  acts like a scaling factor, dilating or shrinking the estimated support, as the subtracting a constant in a frequency domain only affects the value at the origin in the spatial domain. The results of the correlation analysis for the connectivity kernel are shown in Figure~\ref{fig:KernelWidthEstimationExperiment} where each subplot is the normalised  estimate for varying synaptic time constants while holding the observation noise fixed. Note  for clarification the amplitude at the origin was negated when the chosen values of $\tau$ and $\sigma^2_{\varepsilon}$ resulted in the negative values at the origin. The result suggested that a support of approximately $\left[-5,5 \right] $ is a reasonable choice for the spatial extent of the connectivity kernel.

The estimated supports were then employed in the EM algorithm described in \S\ref{subsec:EM-EMAlgorithm} to estimate the smoothed states $\hat{\mathbf x}_t^b$, the connectivity kernel weights $\boldsymbol\theta$, the synaptic dynamic $\xi$, and the parameters $\sigma_{\epsilon}^2$ and $\sigma^2_d$. A  square grid of $n_{\theta}=25$ (5 by 5) basis functions with $2.25~mm$ spacing were used to cover the estimated kernel support where $\sigma_{\psi}^2=3.5$. This is shown in Figure.\ref{fig:KernelWidthEstimation} by blue circles, each of which representing full width at half maximum of the connectivity kernel basis functions. The depicted support was obtained by normalising and thresholding for values less than 0.01.  The red circle shows the estimated support which encompasses the non-zero values (black area) after thresholding. 

@@

Figure~\ref{fig:EM-Histogram} depicts the histogram of the non-kernel parameter estimates from 100 simulations where the true parameter value is shown by the solid line and the mean parameter estimate is shown by the dotted line. These histograms indicate small bias in the parameter estimates with small standard deviations of 0.004, 0.0013 and 0.0012 for Figure~\ref{fig:EM-Histogram}(a), (b) and (c) respectively.

@@

Examples of the neural field reconstruction are given in Figure~\ref{fig:FieldEstimates}.  The figure shows snapshots at three different time instants of the true and estimated neural fields. In each subplot neural activity is plotted against spatial index in 1-D to provide a better comparison between the true and estimated underlying fields. 

@@

Plot of the lower bound on the log-likelihood function versus iterations of the EM is shown in Figure~\ref{fig:LowerBound}. In general, for the non-linear dynamical systems the actual lower bound cannot be calculated analytically, therefore an approximate of the lower bound obtained by URTSS is shown. The mean change in the lower bound in consecutive iterations dropped below 0.009 percent after 15 iterations, confirming
that the algorithm converged to a steady value.

@@

To further illustrate the convergence of the algorithm the parameter estimates averaged over 100 realisation is also shown in Figure~\ref{fig:EM-ParameterConvergence}, demonstrating the convergence of the parameter estimates.

@@

The correlation analysis specifies approximate supports for both the disturbance covariance function and connectivity kernel. Although the unknown parameters $\xi$ and $\sigma_{\varepsilon}^2$ can be set as described in \S\ref{subsec:CorrAnal}, the chosen values influence the exact shape of the kernel significantly, the analysis also assumes the sigmoidal firing rate in the active region, as a result  the kernel and disturbance amplitudes cannot be inferred only through the correlation analysis. This step might be considered as a prerequisite for the EM algorithm described in \S\ref{subsec:EM-EMAlgorithm} where the approximated supports are used to estimate the gain of the kernel and field disturbance variance. Knowing the spatial extent of the connectivity kernel greatly improves the speed of the estimation algorithm as the number of basis functions for the kernel decomposition is significantly reduced  compared to the case where the basis functions are arranged across the entire spatial domain. This also improves the uncertainty of the estimates as lower number of unknown parameters need to be estimated. The approximated support of the disturbance covariance function facilitates the application of the EM algorithm. This way the disturbance covariance matrix can be decomposed into two parts: an unknown scalar and a constant matrix depending only on the inferred spatial support. Therefore, the estimation problem of the disturbance covariance matrix breaks down to the estimation of a single scalar parameter, improving the accuracy and the convergence property of the algorithm.

The additional Taylor approximation step introduced in the E-step of the algorithm resembles the extended Kalman filtering \cite{Jazwinski1970,Chui2009}. EKF approximates a stationary non-linear dynamic system with a linear non-stationary one using first order Taylor expansion around the state estimates. This might imply that the extended Kalman smoother is a better candidate (in terms of implementation) for the state inference in the proposed algorithm. However, extended Kalman smoother (EKS) requires re-linearising about the backwards estimates which in turn necessitates the non-linear mapping to be uniquely invertible. Although this is indeed the case in the neural field equation employed in this work it will impose more computational complexity in the state estimation. In addition, URTSS provides better state approximation compared to the EKS \cite{Haykin2001} and therefore the constructed lower bound will be more accurate, resulting in a better parameter estimation.

While EKS has been  applied in the EM algorithm for learning non-linear models \cite{Ghahramani1999} this is the first work (to the best author's knowledge) where the URTSS is used in this context. However, as in the case of the EKS the stability of the algorithm cannot be ensured. This is because the use of EKS or URTSS represents an approximation to the exact E-step. As a result the maximisation of the approximated lower bound might not necessairly increase the log-likelihood function. Nevertheless, it is envisaged that the use of URTSS will provide a more stable algorithm due to its superior performance.

Typically a sequence of likelihood values converge to some stationary points i.e. global (local) maximum or a saddle point. If the sequence of EM is trapped in a saddle point a small random perturbation diverges the algorithm from such a stationary value \cite{McLachlan1997}. In general convergence of the EM algorithm to any stationary points either global (local) maxima or saddle points depends on the initialisation. Here a bounded sequence of the state vectors were used to initialise the algorithm which gives a satisfactory convergence and good state and parameter estimations.

Since the non-linearity is a sigmoid function first-order Taylor linearisation provides a good approximation to the lower bound of the likelihood function, however in general settings with more complicated non-linear behaviours higher order terms might need to be included, this however introduces intractable computational complexity.
\section{Conclusion}
In this paper, a method has been developed to estimate a spatially continuous neural field model from electrophysiological data where the kernel spatial support and the field disturbance characteristics were unknown to the estimator.  It was shown that by choosing the synaptic parameter and the observation noise variance within a plausible range these unknown quantities can be identified. This improves the complexity and the accuracy of the EM algorithm by reducing the number of unknown variables in the estimation process. In addition to the connectivity kernel weights, synaptic parameter, observation noise and disturbance variances were also identified. The empirical distributions of the estimated non-kernel parameters showed small bias with very smalll standard deviations.

>>>>

\begin{equation}\label{eq:AnalyticSupportCalculation}

@@

This section demonstrates the identification of the proposed IDE model using simulated data. The observation noise and the field disturbance were unknown to the estimation algorithm and no assumption were made on the shape of the connectivity kernel. Data was generated employing state-space equation (\ref{eq:LS-finalformstatespacemodel}-\ref{eq:LS-QmatrixForSigmapoints}) using a square grid of $n_x= 9 \times 9$ equally spaced ($2.5$~mm) basis functions with $\sigma_{\phi}^2=1.58$~mm$^2$, assuming the observed field can be fully described by the decomposed IDE model. 100 realizations of 500 ms of data were generated and the first 100 ms was discarded allowing the model's dynamic to reach the steady state. The observation noise variance was set to 0.2. All the other parameters for the simulation were chosen as that of Chapter~\ref{LSChapter} (see Table~\ref{tab:LS-Parameters}).

\subsection{Correlation analysis}
The correlation analysis described in \S\ref{subsec:CorrAnal} were applied to the generated data in order to approximate the unknown supports of both the connectivity kernel and the disturbance covariance function. In doing so, the unknown parameters $\sigma_{\epsilon}^2$ and $\xi$ are required and should be guessed beforehand.  To form a more accurate initial guess of the unknown parameters and explore the effect of the chosen initial guess on the estimated supports several tests were performed. First consider  \eqref{eq:EM-MMGFourier2} for estimating the disturbance covariance function. As mentioned before the power spectral densities are non-negative, therefore the minimum value of $\mathcal F \{\mathbf E\left[ R_{y_ty_t}\right] \}$ and $\mathcal F \{\mathbf E\left[ R_{y_{t+1}y_{t+1}}\right]\}$ should be chosen as the higher limit for the initial guess of $\sigma^2_{\varepsilon}$. The minimum value of the power spectral densities was found to be 0.23, Figure~\ref{fig:DisturbanceSupportExperiment} shows the support estimates for varying  $\sigma^2_{\varepsilon}$ around the calculated limit. This figure shows that for the values less than 0.23 the estimated support is close to the true one (see Figure~\ref{fig:BinaryDisturbanceWidthEstimation}(b)) while for the values above 0.23 the shape of the disturbance covariance function is distorted. The estimated support of the covariance function convolved twice with the sensor kernel for  $\sigma^2_{\varepsilon}<$0.23 is shown in Figure~\ref{fig:BinaryDisturbanceWidthEstimation} (a), where the results were normalised and thresholded for values less than 0.01. By fitting a Gaussian basis function to the result, and under the assumption that the width of the sensor, $\sigma_m^2$, was known the width of the disturbance covariance function, $\sigma_{\gamma}^2$, was analytically calculated  using \eqref{eq:AnalyticSupportCalculation}. Alternatively deconvolution can be performed numerically to directly find the support as shown in Figure~\ref{fig:DisturbanceWidthEstimationCrossSection}. The effect of the observation noise variance on the connectivity kernel support is very much like that of the disturbance covariance function, where an initial guess can be made in a similar manner. To study the influence of the synaptic parameter, $\xi=1-\frac{ T_s}{\tau}$, on the resultant support of the connectivity kernel different values of the synaptic time constant, $\tau$, in a physiologically plausible range were considered, while noting that the chosen values should be higher than the sampling time, $T_s$. In fact synaptic parameter in \eqref{eq:EM-Fourier_TF_of_Kernel}  acts as a scaling factor --- dilating or shrinking the estimated support --- since subtracting a constant in the frequency domain only affects the value at the origin in the spatial domain. The results of the correlation analysis for the connectivity kernel are shown in Figure~\ref{fig:KernelWidthEstimationExperiment} where each subplot is the normalised kernel estimate for varying synaptic time constants while holding the observation noise fixed. Note for clarification the amplitude at the origin was negated when the chosen values of $\tau$ and $\sigma^2_{\varepsilon}$ resulted in a negative value at the origin.  The result suggested that a support of approximately $\left[-5,5 \right] $ is a reasonable choice for the spatial extent of the connectivity kernel. 

@@

\subsection{State and parameter estimation}
The estimated supports were employed in the EM algorithm described in \S\ref{sec:EstimationAlgorithm} to estimate the smoothed states $\hat{\mathbf x}_t^b$, the connectivity kernel weights $\boldsymbol\theta$, the synaptic dynamic $\xi$, and the parameters $\sigma_{\epsilon}^2$ and $\sigma^2_d$. A  square grid of $n_{\theta}=25$ (5 by 5) basis functions with $2.25$~mm spacing were used to cover the estimated kernel support where $\sigma_{\psi}^2=3.5$~mm$^2$. This is shown in Figure.\ref{fig:KernelWidthEstimation} by blue circles, each of which representing full width at half maximum of the connectivity kernel basis functions. The depicted support was obtained by normalising and thresholding the result for the  values less than 0.01.  The red circle shows the estimated support which encompasses the non-zero values (black area) after thresholding. 

@@

Figure~\ref{fig:EM-Histogram} depicts the histogram of the non-kernel parameter estimates from 100 simulations where the true parameter values are shown by the solid lines and the mean parameter estimates are shown by the dotted lines. These histograms indicate small bias in the parameter estimates with small standard deviations of 0.004, 0.0013 and 0.0012 for Figure~\ref{fig:EM-Histogram}(a), (b) and (c) respectively.

@@

Examples of the neural field reconstruction are given in Figure~\ref{fig:FieldEstimates}. The figure shows snapshots of the true and estimated neural fields at three different time instants. In each subplot the neural activity is plotted against the spatial index in one dimension to provide a better comparison between the true and the estimated underlying fields. 

@@

Plot of the lower bound on the log-likelihood function versus iterations of the EM is shown in Figure~\ref{fig:LowerBound}. In general, for non-linear dynamical systems the actual lower bound cannot be calculated analytically, therefore an approximate of the lower bound obtained by the URTSS is shown. The mean change in the lower bound in consecutive iterations dropped below 0.009 percent after 15 iterations, confirming that the algorithm converged to a steady value.

@@

To further investigate the convergence property of the algorithm the parameter estimates averaged over 100 realisations are also shown in Figure~\ref{fig:EM-ParameterConvergence}, demonstrating the convergence of the EM algorithm.

@@

The correlation analysis specifies approximate supports for both the disturbance covariance function and the connectivity kernel. Although the unknown parameters $\xi$ and $\sigma_{\varepsilon}^2$ can be set as described in \S\ref{subsec:CorrAnal}, the chosen values influence the exact shape of the kernel significantly. The analysis also assumes the sigmoidal firing rate in the active region and as a result  the kernel and disturbance amplitudes cannot be inferred only through the correlation analysis. This step might be considered as a prerequisite for the EM algorithm described in \S\ref{sec:EstimationAlgorithm} where the approximated supports are used to estimate the gain of the kernel and the field disturbance variance. Knowing the spatial extent of the connectivity kernel greatly improves the speed of the estimation algorithm as the number of basis functions for the kernel decomposition is significantly reduced  compared to the case where the basis functions are arranged across the entire spatial domain. This also improves the uncertainty of the estimates as a lower number of unknown parameters need to be estimated. The approximated support of the disturbance covariance function facilitates the application of the EM algorithm. This way the disturbance covariance matrix can be decomposed into two parts: an unknown scalar and a constant matrix depending only on the inferred spatial support. Therefore, the estimation problem of the disturbance covariance matrix breaks down to the estimation of a single scalar parameter, improving the accuracy and the convergence property of the algorithm.

The additional Taylor approximation step introduced in the E-step of the algorithm resembles the extended Kalman filtering \cite{Jazwinski1970,Chui2009}. The EKF approximates a stationary non-linear dynamic system with a linear non-stationary one using the first order Taylor expansion around the state estimates. This might imply that the extended Kalman smoother (EKS) is a better candidate --- in terms of implementation --- for the state inference in the proposed algorithm. However, the EKS requires re-linearising about the backward estimates which in turn necessitates the non-linear mapping to be uniquely invertible. Although this is indeed the case in the neural field equation employed in this work it will impose more computational complexity in the state estimation. In addition, the URTSS provides better state estimations compared to the EKS \cite{Haykin2001} and therefore the constructed lower bound will be more accurate, resulting in a better parameter estimation.

While the EKS has been  applied in the EM algorithm for learning non-linear models \cite{Ghahramani1999} this is the first work (to the best author's knowledge) where the URTSS is used in this context. However, as in the case of the EKS the stability of the algorithm cannot be ensured. This is because the use of the EKS or URTSS represents an approximation to the exact E-step. As a result the maximisation of the approximated lower bound might not necessarily increase the log-likelihood function. Nevertheless, it is envisaged that the use of the URTSS will provide a more stable algorithm due to its superior performance.

Typically a sequence of likelihood values converge to some stationary points i.e. global (local) maximum or a saddle point. If the sequence of the EM is trapped in a saddle point a small random perturbation diverges the algorithm from such a stationary value \cite{McLachlan1997}. In general convergence of the EM algorithm to any stationary points either global (local) maxima or saddle points depends on the initialisation. Here a bounded sequence of the state vectors were used to initialise the algorithm which gives a satisfactory convergence and good state and parameter estimations.

Since the non-linearity is a sigmoid function first-order Taylor linearisation provides a good approximation to the lower bound of the likelihood function, however in general settings with more complicated non-linear behaviour higher order terms might need to be included, this however introduces intractable computational complexity.
\section{Conclusion}
In this chapter, a method has been developed to estimate a spatially continuous neural field model from electrophysiological data, where the kernel spatial support and the field disturbance characteristics were unknown to the estimator.  It was shown that by choosing the synaptic parameter and the observation noise variance within a plausible range these unknown quantities can be identified. This improves the complexity and the accuracy of the EM algorithm by reducing the number of unknown variables in the estimation process. In addition to the connectivity kernel weights and the synaptic parameter, the observation noise and disturbance variances were also identified. The empirical distributions of the estimated non-kernel parameters showed small bias with very small standard deviations.
